---
title: "Assignment 7: High Frequency Data"
author: "Ethan Ready"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Hydrologic Data Analysis on high frequency data

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single pdf file.
5. After Knitting, submit the completed exercise (pdf file) to the dropbox in Sakai. Add your last name into the file name (e.g., "A07_Chamberlin.pdf") prior to submission.

The completed exercise is due on 16 October 2019 at 9:00 am.

## Setup

1. Verify your working directory is set to the R project file, 
2. Load the StreamPULSE, streamMetabolizer and tidyverse packages. 
3. Set your ggplot theme (can be theme_classic or something else)


```{r setup}
getwd()

library(devtools)
library(tidyverse)

packages <- c(
  "tidyverse", 
  "StreamPULSE", 
  "streamMetabolizer"
  )
invisible(
  suppressPackageStartupMessages(
    lapply(packages, library, character.only = TRUE)
    )
  ) 

theme_set(theme_classic(base_size = 12))

```


4. Download data from the Stream Pulse portal using `request_data()` for the Kansas River, ("KS_KANSASR"). Download the discharge (`Discharge_m3s`), disolved oxygen (`DO_mgL`) and nitrate data (`Nitrate_mgL`) for the entire period of record

```{r}
Kansas.Dat<-request_data(
  sitecode = "KS_KANSASR",
  variables = c("Discharge_m3s", "DO_mgL", "Nitrate_mgL"),
)

```


5. Reformat the data into one dataframe with columns DateTime_UTC, DateTime_Solar (using `convert_UTC_to_solartime()`), SiteName, DO_mgL, Discharge_m3s, and Nitrate_mgL.
```{r Datadownload}
Kansas.df<-select(Kansas.Dat[[1]], DateTime_UTC, value, variable)
  
Kansas.df<-spread(Kansas.df, value = value, key = variable)
  
Kansas.df<- mutate(Kansas.df, DateTime_Solar = convert_UTC_to_solartime(DateTime_UTC, Kansas.Dat[[2]]$lon))


```

6. Plot each of the 3 variables against solar time for the period of record

```{r}
ggplot(Kansas.df, aes(x=DateTime_Solar))+
  geom_point(aes(y=DO_mgL))+
  labs(x="Solar Date and Time", y= "Dissolved Oxygen (mg/L)")

ggplot(Kansas.df, aes(x=DateTime_Solar))+
  geom_point(aes(y=Nitrate_mgL))+
  labs(x="Solar Date and Time", y= "Nitrate (mg/L)")
  
ggplot(Kansas.df, aes(x=DateTime_Solar))+  
  geom_point(aes(y=Discharge_m3s))+
  labs(x="Solar Date and Time", y= "Discharge (m3/s)")
```

7. How will you address gaps in these dataseries?

> I will omit all rows with missing data

8. How does the daily amplitude of oxygen concentration swings change over the season? What might cause this?

> It's higher in months with more sunlight because more primary production is happening.

## Baseflow separation
9. Use the `EcoHydRology::BaseflowSeparation()` function to partition discharge into baseflow and quickflow, and calculate how much water was exported as baseflow and quickflow for this time period. Use the DateTime_UTC column as your timestamps in this analysis.

```{r}
Kansas.no.na<-na.omit(Kansas.df)

Kansas.baseflow<-EcoHydRology::BaseflowSeparation(Kansas.no.na$Discharge_m3s,
                                    filter_parameter = 0.925, 
                                    passes = 3)

Kansas.new<-cbind(Kansas.no.na, Kansas.baseflow)

Kansas.flow.diff<-mutate(Kansas.new, timestep = c(diff(as.numeric(DateTime_UTC)),NA_real_),
         baseflow.gross = bt*timestep,
         quickflow.gross = qft*timestep)

Kansas.flow.diff<-summarize(Kansas.flow.diff, BF.total = sum(baseflow.gross, na.rm = TRUE),
            QF.total = sum(quickflow.gross, na.rm = TRUE),
            flow.total = BF.total+QF.total,
            BF.percent = BF.total/flow.total,
            QF.percent = QF.total/flow.total)

  
```


The `package::function()` notation being asked here is a way to call a function without loading the library. Sometimes the EcoHydRology package can mask tidyverse functions like pipes, which will cause problems for knitting. In your script, instead of just typing `BaseflowSeparation()`, you will need to include the package and two colons as well.

10. Create a ggplot showing total flow, baseflow, and quickflow together. 


```{r}
ggplot(Kansas.new, aes(x = DateTime_UTC))+
  geom_line(aes(y = Discharge_m3s))+
  geom_line(aes(y = bt), color = "pink")+
  geom_line(aes(y = qft), color = "steelblue")
```


11. What percentage of total water exported left as baseflow and quickflow from the Kansas River over this time period?

> Baseflow: 95.7%, Quickflow: 4.3%

12. This is a much larger river and watershed than the 2 we investigated in class. How does the size of the watershed impact how flow is partitioned into quickflow and baseflow? 

> A larger watershed probably has more baseflow and isn't as volatile as the Eno or other smaller creeks. A big watershed means that a storm probably doesn't hit the whole area at once, allowing a tributary to stabilize others that may be hit by quickflow during a big storm.

13. The site we are looking at is also further down in its river network (i.e. instead of being a headwater stream, this river has multiple tributaries that flow into it). How does this impact your interpretation of your results?

> Each tributary will react to different regional storms, minimizing the impact that one storm has on a watershed.

## Chemical Hysteresis

14. Create a ggplot of flow vs. nitrate for the large storm in May (~May 1 - May 20). Use color to represent Date and Time.

```{r}
Kansas.storm<-
  filter(Kansas.no.na, DateTime_UTC> "2018-05-1" & DateTime_UTC < "2018-05-31")

ggplot(Kansas.storm, aes(x = Nitrate_mgL, y = Discharge_m3s, color = DateTime_Solar))+
  geom_point()
```

15. Does this storm show clockwise or counterclockwise hysteresis? Was this storm a flushing or diluting storm?

> I'm confused, because in class we plotted discharge on the x and nitrate on the y, but this question asked for them flipped. In class, we discussed that a countrclockwise loop was a diluting storm because it suggests that quickflow has lower nutrient levels, which is what this storm appears to be. If its actually supposed to be discharge on y then my answer would be flipped.

16. What does this mean for how nitrate gets into the river from the watershed?

> This would mean that more nitrate gets into the watershed from baseflow.

## Reflection
17. What are 2-3 conclusions or summary points about high frequency data you learned through your analysis?

> EcoHydRology masks pipes. It's important which axis you plot things on.

18. What data, visualizations, and/or models supported your conclusions from 17?

> The Hysterysis plot

19. Did hands-on data analysis impact your learning about high frequency data relative to a theory-based lesson? If so, how?

> yes I understood how frustrating R can be

20.	How did the real-world data compare with your expectations from theory?

> It made sense once I looked at it, but I didn't really have many expectations going in.
